The Way Forward in Parallelizing Dynamic Languages

### 2014-05-01 talk/icooolps2014/position-paper.tex
TM draft

\subsection{Why is there a GIL?}
The GIL is a very simple synchronization mechanism for supporting
multi-threading in the interpreter. The basic guarantee is that the
GIL may only be released in-between bytecode instructions. The
interpreter can thus rely on complete isolation and atomicity of these
instructions. As a consequence, applications can rely on certain
operations to be atomic. While this is probably not a good idea,
it is used in practice. A solution replacing the GIL should therefore
uphold these guarantees, while preferably also be as easily
implementable as a GIL for the interpreter.
[xxx mention that the interpreter is typically very large and maintained
by open-source communities]

@@

\subsubsection{Transactional Memory}
\paragraph{HTM}

- false-sharing on cache-line level\\
- limited capacity (caches, undocumented)\\
- random aborts (haswell)\\
- generally: transaction-length limited (no atomic blocks)

@@

- overhead (100-1000\%) (barrier reference resolution, kills performance on low \#cpu)
(FastLane: low overhead, not much gain)\\
- unlimited transaction length (easy atomic blocks)

@@

%                       Revision History
%                       -------- -------
%  Date         Person  Ver.    Change
%  ----         ------  ----    ------

%  2013.06.29   TU      0.1--4  comments on permission/copyright notices

>>>>

\subsection{Why is there a GIL?}
The GIL is a very simple synchronization mechanism for supporting
multi-threading in the interpreter. The basic guarantee is that the
GIL may only be released in-between bytecode instructions. The
interpreter can thus rely on complete isolation and atomicity of these
instructions. Additionally, it provides the application with a
sequential consistency model. As a consequence, applications can rely
on certain operations to be atomic and that they will always be
executed in the order in which they appear in the code. While
depending on this may not always be a good idea, it is done in
practice. A solution replacing the GIL should therefore uphold these
guarantees, while preferably also be as easily implementable as a GIL
for the interpreter.
[xxx mention that the interpreter is typically very large and maintained
by open-source communities]

@@

\subsubsection{Transactional Memory}
Transactional memory (TM) can be used as a direct replacement for a
single global lock. Transactions provide the same atomicity and
isolation guarantees as the GIL provides for the execution of bytecode
instructions. So instead of acquiring and releasing the GIL between
these instructions, this approach runs the protected instructions
inside transactions.

TM can be implemented in software (STM) or in hardware (HTM. There are
also some hybrid approaches that combine the two. We count these
hybrid approaches as STM, since they usually provide the same
capabilities as software-only approaches but with different
performance characteristics. We will now first look at HTM that
recently gained a lot of popularity by its introduction in common
desktop CPUs from Intel (Haswell generation).

\paragraph{HTM}

HTM provides us with transactions like any TM system does. It can
be used as a direct replacement for the GIL. However, as is common
with hardware-only solutions, there are quite a few limitations
that can not be lifted easily. For this comparison, we look at
the implementation of Intel in recent Haswell generation CPUs.

HTM in these CPUs works on the level of caches. This has a few
consequences like false-sharing on the cache-line level, and most
importantly it limits the amount of memory that can be accessed within
a transaction. This transaction-length limitation makes it necessary
to have a fallback in place in case this limit is reached. In recent
attempts, the usual fallback is the GIL (XXX: cite). The current
generation of HTM hits this limit very often for our use case (XXX:
cite ruby GIL paper) and therefore does not parallelize that well.

The performance of HTM is pretty good (XXX: cite again...) as it does
not introduce much overhead. And it can transparently parallelize
existing applications to some degree. The implementation is very
straight-forward because it directly replaces the GIL in a central
place. HTM is also directly compatible with any external library that
needs to be integrated and synchronized for use in multiple
threads. The one thing that is missing is support for a better
synchronization mechanism for the application. It is not possible
in general to expose the hardware-transactions to the application
in the form of atomic blocks because that would require much
longer transactions.

%% - false-sharing on cache-line level\\
%% - limited capacity (caches, undocumented)\\
%% - random aborts (haswell)\\
%% - generally: transaction-length limited (no atomic blocks)

@@

STM provides all the same benefits as HTM except for its performance.
It is not unusual for the overhead introduced by STM to be between
100\% to even 1000\%. While STM systems often scale very well to a big
number of threads and eventually overtake the single-threaded
execution, they often provide no benefits at all for low numbers of
threads (1-8). There are some attempts (XXX: cite fastlane) that can
reduce the overhead a lot, but also scale very badly so that their
benefit on more than one thread is little.

However, STM compared to HTM does not suffer from the same restricting
limitations. Transactions can be arbitrarily long.  This makes it
possible to actually expose transactions to the application in the
form of atomic blocks. This is the only approach that enables a better
synchronization mechanism than locks for applications \emph{and} still
parallelizes when using it. We think this is a very important point
because it not only gives dynamic languages the ability to parallelize
(already commonplace in most other languages), but also pushes
parallel programming forward. Together with sequential consistency it
provides a lot of simplification for parallel applications.

%% - overhead (100-1000\%) (barrier reference resolution, kills performance on low \#cpu)
%% (FastLane: low overhead, not much gain)\\
%% - unlimited transaction length (easy atomic blocks)

@@
